The core transcriptome of mammalian placentas and the divergence of expression with placental shape

### 2016-11-18 manuscript/manuscript.Rnw
fix go analyses ref

@@

Using a mixed-effects linear model to account for repeated testing of
\human and \mouse, we identified multiple genes which are
differentially expressed in the \human lineage in comparison to all
other lineages (\cref{tab:human_lineage_specific}). These genes
include multiple genes which are thought to be involved in
pre-eclampsia. Notably, \textit{CRH}
(FDR=\Sexpr{hum.all.lin.spec["CRH",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["CRH",`Fold Change`]}, Supplemental
\cref{fig:crh_evolution_tree}), has been shown to be more highly
expressed in placentas from pre-eclamptic pregnancies
\cite{Petsas.ea2012:Aberrantexpressionofcorticotropin-releasing,Paiva.ea2011:MeasurementofmRNAtranscripts},
where it is hypothesized to activate decidual macrophages. Likewise,
\textit{KISS1} (\Sexpr{hum.all.lin.spec["KISS1",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["KISS1",`Fold Change`]}, Supplemental
\cref{fig:kiss1_evolution_tree}) also shows increased expression in
pre-eclamptic placentas
\cite{Vazquez-Alaniz.ea2011:Comparativeexpressionprofilesfor}, and is
involved in modulating trophoblast invasion. Two other genes,
\textit{PAPPA} (\Sexpr{hum.all.lin.spec["PAPPA",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["PAPPA",`Fold Change`]} Supplemental
\cref{fig:pappa_evolution_tree}) and \textit{ADAM12}
(\Sexpr{hum.all.lin.spec["ADAM12",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["ADAM12",`Fold Change`]}, Supplemental
\cref{fig:adam12_evolution_tree}) are also differentially expressed
but are negatively correlated with pre-eclampsia
\cite{Spencer.ea2008:ADAM12sinmaternalserum}.

@@

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all species studied and was
not a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>=1 & 
`Min. Exp. in` == '14 of 14',max(num_genes)]} %
genes in the core transcriptome. %
In contrast to the shape of the number of genes included in the core
placenta transcriptome, the set of \human housekeeping genes expressed
all placentas is flatter (Supplemental
\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
of the total fraction of genes. However, there is still a large
difference between the set of housekeeping genes which are expressed
in \human and those which are expressed in all placentas.

>>>>

% \subsubsection{Per-species Gene Set and Pathway Enrichment}

We performed \ac{gsea} on both the top 1\% of transcripts
(Supplemental \cref{tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_spalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_canis_familiaris,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_monodelphis_domestica}) and the top
100 transcripts (Supplemental
\cref{tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_spalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_canis_familiaris,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_monodelphis_domestica}); common processes
include protein transport and viral transcription/expression
(Supplemental \cref{tab:go_terms_all_species}).

@@

Using a mixed-effects linear model to account for repeated testing of
\human and \mouse, we identified multiple genes which are
differentially expressed in the \human lineage in comparison to all
other lineages (\cref{tab:human_lineage_specific}, Supplemental
\cref{tab:human_lineage_specific_all}). These genes include multiple
genes which are thought to be involved in pre-eclampsia. Notably,
\textit{CRH} (FDR=\Sexpr{hum.all.lin.spec["CRH",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["CRH",`Fold Change`]}, Supplemental
\cref{fig:crh_evolution_tree}), has been shown to be more highly
expressed in placentas from pre-eclamptic pregnancies
\cite{Petsas.ea2012:Aberrantexpressionofcorticotropin-releasing,Paiva.ea2011:MeasurementofmRNAtranscripts},
where it is hypothesized to activate decidual macrophages. Likewise,
\textit{KISS1} (\Sexpr{hum.all.lin.spec["KISS1",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["KISS1",`Fold Change`]}, Supplemental
\cref{fig:kiss1_evolution_tree}) also shows increased expression in
pre-eclamptic placentas
\cite{Vazquez-Alaniz.ea2011:Comparativeexpressionprofilesfor}, and is
involved in modulating trophoblast invasion. Two other genes,
\textit{PAPPA} (\Sexpr{hum.all.lin.spec["PAPPA",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["PAPPA",`Fold Change`]} Supplemental
\cref{fig:pappa_evolution_tree}) and \textit{ADAM12}
(\Sexpr{hum.all.lin.spec["ADAM12",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["ADAM12",`Fold Change`]}, Supplemental
\cref{fig:adam12_evolution_tree}) are also differentially expressed
but are negatively correlated with pre-eclampsia
\cite{Spencer.ea2008:ADAM12sinmaternalserum}.

@@

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all species studied and was not
a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}. Genes which
meet this criteria but are housekeeping genes are depicted in
Supplemental \cref{fig:core_placenta_transcriptome_housekeeping}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff
of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>=1 & `Min. Exp. in`
  == '14 of 14',max(num_genes)]} %
genes in the core transcriptome. %
In contrast to the shape of the number of genes included in the core
placenta transcriptome, the set of \human housekeeping genes expressed
all placentas is flatter (Supplemental
\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
of the total fraction of genes. However, there is still a large
difference between the set of housekeeping genes which are expressed
in \human and those which are expressed in all placentas.
